Template for indicator documentation

### 2023-09-11 functional_plant_indicators_wetland.Rmd
RenderBook and remove sysDate. Fixed some typos in gjengoing and image widths

# Functional plant indicators, wetland ecosystems (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}

@@

_Author and date:_
Joachim Töpper

@@

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for wetland ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the moisture signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

## About the underlying data
In the 'functional plant indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:

@@

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which as of today comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

## Uncertainties
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.

@@

```{r, echo = T, eval = F}
colnames(NiN.wetland)
# 1st column is the species
# 2nd-46th column is the abundances of sp in different ecosystem types
# 47th-50th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
# each sample re-samples 1/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
                          iter=1000,obl=1,rat=1/3,var.abun=T)

@@

```{r, include = F}
# Data from cache
wetland.ref.cov<-readRDS(paste0(here::here(), "/data/cache/wetland.ref.cov.RDS"))

#load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
```

@@

Once test data (ANO) and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper condition indicators for each functional plant indicator. An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

@@

In order to visualize the results we need to rearrange the results-objects from wide to long format (note that there is both a lower and an upper condition indicator for each of the functional plant indicators).

@@

# make long version of the scaled value part
res.wetland <-
  res.wetland %>% 
  pivot_longer(
    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
    names_to = "fp_ind",
    values_to = "scaled_value",
    values_drop_na = FALSE
  )

# add original values as well
res.wetland <- 
  res.wetland %>% add_column(original = results.wetland[['original']] %>% 
                               pivot_longer(
                                 cols =  c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
                                 names_to = NULL,
                                 values_to = "original",
                                 values_drop_na = FALSE
                               ) %>%
                               pull(original)
  )

@@

# making the plot
ggplot(res.wetland[!is.na(res.wetland$scaled_value),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Main ecosystem type") + 
  ylab("Scaled indicator value") 
```
The functional signature from the plant communities in wetland ecosystems looks largely good, except for V2 (swamp forests), which show rather low index scores for the lower pH and Nitrogen indicators, i.e. suggesting too acidic conditions and too low nitrogen availability.

We can look at V2 specifically to see if this is the case for all sub-types.

```{r}
# making the plot for V2 only
ggplot(res.wetland[res.wetland$hovedtype_rute=="V2",], aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
  #  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Main ecosystem type") + 
  ylab("Scaled indicator value") 
```
The deviations towards low pH and low nitrogen are mainly an issue for V2-C1, limestone-poor swamp forests.

@@

### Regions - maps and statistics
Alternatively we can calculate and show the region-wise means and their related standard errors. But note that calculating a simple mean would be inappropriate for these data. This is because:
(i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
(ii) the ANO dataset has a nested structure

Therefore, we need to (i) use a beta-model, that (ii) can account for the nested structure of the data.
Here, we apply the following function using either a glmmTMB null-model with a beta-distribution, logit link, and the nesting as a random intercept, or a simple betareg null-model with logit link if the nesting is not extensive enough for a mixed model.
```{r}
library(betareg)
library(glmmTMB)

expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale

# the function performs a glmmTMB if there's >= 5 random levels in the nesting structure
# if that is not the case, then the function performs a betareg if theres >= 2 observations
# if that is not the case either, then the function returns the value of the single observation with NA standard error

indmean.beta <- function(df) {

  st_geometry(df) <- NULL
  colnames(df) <- c("y","ran")

  if ( nrow(df[!is.na(df[,1]),]) >= 2 ) {

    if ( length(unique(df[!is.na(df[,1]),2])) >=5 ) {

      mod1 <- glmmTMB(y ~ 1 +(1|ran), family=beta_family(), data=df)

      return(c(
        expit(summary( mod1 )$coefficients$cond[1]),

        expit( summary( mod1 )$coefficients$cond[1] + 
                 summary( mod1 )$coefficients$cond[2] )-
          expit( summary( mod1 )$coefficients$cond[1] ),

        nrow(df[!is.na(df$y),]),
        summary( mod1 )$coefficients$cond[1],
        summary( mod1 )$coefficients$cond[2]
      ))

    } else {

      mod2 <- betareg(y ~ 1, data=df)

      return(c(
        expit(summary( mod2 )$coefficients$mean[1]),
        expit( summary( mod2 )$coefficients$mean[1] + 
                 summary( mod2 )$coefficients$mean[2] )-
          expit( summary( mod2 )$coefficients$mean[1] ),
        nrow(df[!is.na(df$y),]),
        summary( mod2 )$coefficients$mean[1],
        summary( mod2 )$coefficients$mean[2]
      ))

  } else {

    return(c(df$y,NA,1,NA,NA))

@@

```{r}

regnor <- regnor %>%
  mutate(
    pH1.reg.mean = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[1],
                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[1],
                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[1],
                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[1],
                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[1]
    ),
    pH1.reg.se = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[2],
                               indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[2],
                               indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[2],
                               indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[2],
                               indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[2]
    ),
    pH1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH1),]),
                              nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$pH1),]),
                              nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$pH1),]),
                              nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH1),]),
                              nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),])
    )
  )

## scaled value maps for pH1 (lower indicator)
# mean
tm_shape(regnor) +
  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
  tm_text("pH1.reg.n",col="black",bg.color="grey")
```
Mean index value by region for the lower pH indicator (i.e. index shows deviations towards a more acidic environment). Numbers in grey fields show the number of observations in the respective region.

```{r}
# se
tm_shape(regnor) +
  tm_polygons(col="pH1.reg.se", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("pH1.reg.n",col="black",bg.color="grey")

```
Standard error to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.

### unscaled values vs. reference
We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation. Since pH and nitrogen show some deviation of the lower pH indicator we exemplify this with these indicators for unscaled values.

```{r}
summary(res.wetland$kartleggingsenhet_1m2)
length(unique(res.wetland$kartleggingsenhet_1m2))
# 16 NiN-types to plot
colnames(wetland.ref.cov[['Soil_reaction_pH']])

### pH
par(mfrow=c(2,4))
## V1s
# V1-C-1
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C1',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
       col="red")

# V1-C-2
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C2',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
       col="red")

# V1-C-3
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C3',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
       col="red")

# V1-C-4
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C4',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
       col="red")

# V1-C-5
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C5")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C5',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
       col="red")

# V1-C-6
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C6',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
       col="red")

# V1-C-7
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C7',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
       col="red")

# V1-C-8
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C8',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
       col="red")

par(mfrow=c(2,4))
## V2s
# V2-C-1
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C1',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
       col="red")

# V2-C-2
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C2',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
       col="red")

# V2-C-3
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C3',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
       col="red")

## V3s
# V3-C-1
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V3-C1',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
       col="red")

# V3-C-2
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C2")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V3-C2',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
       col="red")

## V4s
# V4-C-2
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V4-C2")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V4-C2',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
       col="red")

## V8s
# V8-C-1
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C1")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V8-C1',xlab='pH value')
#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
#       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
       col="red")

# V8-C-2
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C2")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V8-C2',xlab='pH value')
#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
#       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
       col="red")
legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))

>>>>

# Functional plant indicators (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}

@@

_Author and date:_

Joachim Töpper

@@

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for wetland ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021).

## About the underlying data
In the 'plant functional indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:

@@

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

## Uncertainties
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.

@@

```{r, echo = T, eval = F}
colnames(NiN.wetland)
# 1st column is the species
# 2nd-46th column is the abundances of sp in different ecosystem types
# 47th-50th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
# each sample re-samples 2/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
                          iter=1000,obl=1,rat=2/3,var.abun=T)

@@

```{r, include = F}
# Data from cache
wetland.ref.cov<-readRDS(paste0(here::here(), "/data/cache/wetland.ref.cov.RDS"))
#load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
```

@@

Once test data (ANO) and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type. An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

@@

### Simple summary statistics
We can calculate simple summary statistics like means, standard deviations, and number of observations (note that there is both a lower and an upper indicator for each of the four plant functional indicators)

@@

# make long version of the scaled value part
res.wetland <-
  res.wetland %>% 
  pivot_longer(
    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
    names_to = "fp_ind",
    values_to = "scaled_value",
    values_drop_na = TRUE
  )

# summarizing the indicator scores
res.wetland %>%
  group_by(fp_ind) %>%
  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value))

@@

# making the plot
ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Main ecosystem type") + 
  ylab("Scaled indicator value") 
```

@@

### Regions
Alternatively we can calculate and show the region-wise means and standard deviations
```{r}

regnor <- regnor %>%
  mutate(
    pH1.reg.mean = c(mean(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
                             mean(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
                             mean(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
                             mean(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
                             mean(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
    pH1.reg.sd = c(sd(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
                          sd(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
                          sd(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
                          sd(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T))
    )

## scaled value maps for pH1 (lower indicator)
# mean
tm_shape(regnor) +
  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))

# sd
tm_shape(regnor) +
  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE))

@@
